---
title: 'Phenograss: Survival analysis of germination data'
author: "Jon Yearsley"
date: "26th March 2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup and load data

```{r message=FALSE}
rm(list=ls())

library(survival)
library(survminer)
library(timereg)
library(broom)
library(pander)
library(emmeans)
library(DHARMa)

# Load data
load('germination_data.RData')
```

## Overall visualisation

```{r echo=FALSE}
# Look at distribution of germination times
ggplot(data=subset(germination, Day<30), 
       aes(x=Day, fill=Treatment)) +
  geom_histogram(bins=20) +
  theme_bw() +
  scale_fill_brewer(palette ='Dark2') +
  labs(x='Day post sowing',
         y='Number Germinated',
       title='Germination time distribution')
```

Look at overall numbers that germinated
```{r}
germin_tab = table(germination$Treatment, 
                   germination$germinated)
pander(germin_tab, 
       caption = "Numbers germinating & not germinating within observation period")
```

Test to see if there's an overall difference 
```{r}
pander(chisq.test(germin_tab), 
       caption = "Chi squared test of association on germin_tab")
```

No difference in overall germination rate due to treatment

### Linear modelling of germination probability

Fit logistic model and test for difference between species and treatments

```{r}
m=glm(germinated~Treatment*Variety, data=germination, family='binomial')
```

Do some model validation

```{r}
summary(m)
```
The residual deviance divided by degrees of freedom is about 1, so an indication that there's no overdispersion.


```{r}
# Simulate residuals from the model
val = simulateResiduals(m, n=500, refit=FALSE)
```

Create some validation plots
```{r}
plot(val)
```

```{r}
testResiduals(val)
```


Logistic model looks valid. Fit null models and do some hypothesis testing.

```{r}
m0 = update(m, .~.-Treatment:Variety)

pander(anova(m0,m, test='Chisq'))
```

No evidence of an interaction for total germination proportion


Test for effect of variety
```{r}
m0b = update(m, .~.-Treatment:Variety - Variety)
pander(anova(m0b,m0, test='Chisq'))
```

A strong effect of Variety

Test for effect of Treatement
```{r}
m0c = update(m, .~.-Treatment:Variety - Treatment)
pander(anova(m0c,m0, test='Chisq'))
```

No effect of Treatment


Posthoc test for the effect of Variety and compare each 
```{r}
m_eff = emmeans(m0, specs = 'Variety')
m_posthoc = contrast(m_eff, method='eff')
m_posthoc
```

Plot the posthoc analysis

```{r echo=FALSE}
plot(m_posthoc)
```



## Calculate Kaplan-Meier survival curve

This is a non-parametric approach, and doesn't have the same assumptions as Cox PH model. We use the approach to estimate the survival curve for germination, and to test if Treatment has an effect (averaged across varieties)

The survival curve normally corresponds to the probability of dying. In our case it is the probability of germinating. The hazard rate has a similar interpretation.

### Survival curve for each Variety and Treatment
```{r }
# Effect of Treatment for each Variety
germ_fit = survfit(Surv(Day, germinated) ~ Variety + Treatment, 
                             data=germination)
```

Display median survival times (i.e. germination times) and confidence intervals
```{r}
# Print estimates of time to germination
pander(germ_fit)
```



### Look at survival curves, averaging over varieties

```{r}
# Effect of treatment averaging over variety
germ_fit_treatment = survfit(Surv(Day, germinated) ~ Treatment, 
                             data=germination)
```

Display 

```{r}
# Display results of average time to germination across varieties
pander(germ_fit_treatment)

```

Plot survival curves for Treatments (averaging across varieties)
```{r echo=FALSE}
ggsurvplot(fit=germ_fit_treatment,
           xlab='Days',
           ylab='Proportion of seed not germinated',
           conf.int=TRUE,
           pval=TRUE)
```


#### Hypothesis test

Perform a log-ratio test of difference between survival curves for control - treatment

```{r}
pander(survdiff(Surv(Day, germinated) ~ Treatment, 
         data=germination))
```


## Parametric models of germination


Try to fit a Cox proportional hazards model. This is a parametric model that assumes a constant proportional hazard rate. This assumption must be validated

Fit model with Treatment and Variety. There's no evidence of an interaction between Treatment and Variety
```{r}
# Fit the full model with main effects and interaction
coxph.fit <- coxph(Surv(Day, germinated) ~ Treatment+Variety, 
                   data=germination, 
                   method="breslow")  # Could use efron
```


Same model but treat Variety as a random effect
```{r}
# Fit the full model with main effects and random term for Variety
coxph.rand <- coxph(Surv(Day, germinated) ~ Treatment + frailty(Variety),
                    data=germination)
```


Validate the proportional hazard assumption
```{r}
# Validate hazard function assumption
test.ph <- cox.zph(coxph.fit)
pander(test.ph$table)
```
Looks like the assumption is violated. Meaning that the effect of Treatment and Variety on germination rate varies with time.

Look at some visuals

```{r}
plot(test.ph)
```

Looks like proportional hazard decreases over time for the treatment effect. The effect of Variety seems to be less of an issue.

Look at model with Variety as a random effect
```{r}
test2.ph <- cox.zph(coxph.rand)
pander(test2.ph$table)
```


```{r}
plot(test2.ph)
```

Same message. Looks like proportional hazard decreases over time for the treatment effect

Need to find time varying version of Cox model.





### Effect of variety

Use Cox PH model on data from the control

```{r}
# Fit the full model
coxph.fit_full <- coxph(Surv(Day, germinated) ~ Variety, 
                        data=subset(germination, Treatment=='CON'), 
                        method="breslow")  # Could use efron

# Fit null model testing for Variety on control data
coxph.fit_null <- coxph(Surv(Day, germinated) ~ 1, 
                         data=subset(germination, Treatment=='CON'), 
                         method="breslow")  # Could use efron

# Perform hypothesis test
anova(coxph.fit_full, coxph.fit_null, test='ChiSq')
```


Strong effect of Variety on germination. Plot the hazard ratio for the model with main effect variety
```{r echo=FALSE}
ggforest(coxph.fit_full, data=germination)
```


```{r}
survfit(coxph.fit_full)
```



## Time varying coefficients

### Modify Cox PH model (fit with time bins)

Bin data into 3 bins t<14, 14<t<18, t>18

```{r}
# Create dataframe with time bins
germin2 <- survSplit(Surv(Day, germinated) ~ ., 
                  data= germination, 
                  cut=c(14,18),
                  episode= "tgroup", id="id")
```


```{r}
# Fit model
coxph.fit_full2 <- coxph(Surv(Day, germinated) ~ Treatment:strata(tgroup), 
                        data=germin2, 
                        method="breslow")  # Could use efron
```

```{r}
# Look at mean fitted values
coxph.fit_full2$means
```



### Fit model with time varying coefficients

The hazard rate for Treatment will be allowed to vary with time

```{r eval=TRUE}
m = timecox(Surv(Day, germinated) ~ const(Variety)+Treatment, 
            data=germination, 
            max.time=25, 
            residuals=TRUE)
```


Plot results, and add in null expectation of no variation with time.
```{r echo=FALSE}
plot(m,score=TRUE)
```

```{r}
summary(m)
```


