---
title: "Height_report"
author: "Dana Looschelders"
date: "3 5 2020"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# Height as growth rate significance tests  
 by Dana Looschelders  
#This script performs the following analysis:
  #create four boxplot (one for each treatment) for varieties
  #create boxplot for treatment
  #create boxplot for variety
  #plot histogram of growth rate until first cut
  #test assumptions for parametric tests (discepancy between qqplot and shapiro test)
  #Kruskal-Wallis test
  #posthoc nemenyi test
  #when normality is assumed based on qqplot: bartletts test and t.test
  #boxcox transformation
  

```{r, message=FALSE, echo=FALSE, include=FALSE}
library(tidyverse)
library(MASS)
library(tseries)
library(reshape2)
library(agricolae)
library(PMCMR)
setwd("C:/00_Dana/Uni/Internship/Work/Data Rosemount/")

#load data table with measured height values
height=read.table("Plant Height Data.csv", sep=";", dec=",", header=T)
str(height)
height=Filter(function(x)!all(is.na(x)), height) #remove columns with NA
height=na.omit(height) #remove rows with NA
str(height) #check
height$Treatment=as.character(height$Treatment) #change factor to charactor to remove one empty factor level
height$Treatment=as.factor(height$Treatment) #change back to factor (ugly workaround but I didn't find a faster solution)
str(height)
#calculate growth rate (between cuts)
#biomass cuts:
#1 cut: after week 5
height$growthrate_1cut=(height[,9]-height[,5])/(length(height[,5:9])*7) #calculate dailygrowth rate
mean(height$growthrate_1cut) #mean: 0.5170238
#2 cut: after week 9
height$growthrate_cut2=(height[,13]-height[,10])/(length(height[,10:13])*7) #calculate growth rate
mean(height$growthrate_cut2) #mean: 1.096057
#3 cut: after week 13
height$growthrate_cut3=(height[,17]-height[,14])/(length(height[,14:17])*7) #calculate growth rate
mean(height$growthrate_cut3) #mean: 0.6936331
#global growth rate
height$growthrate_global=rowSums(height[20:22])/length(height[20:22]) #calculate growth rate
mean(height$growthrate_global) #global mean: 0.7689045

#QAQC
any(height[,20:22]<0) #there are some negative growth rates
which(height[,20:22]<0) #find out which ones (236 and 806)
```

# Data Exploration  
```{r, message=FALSE}
#display height with line plot
matplot(t(height[,5:19]), type="l", xlab="weeks", ylab="Plant Height [cm]")

summary(height)

#boxplot
hist(height$growthrate_global)
boxplot(height$growthrate_global~height$Treatment)
boxplot(height$growthrate_global~height$Variety)
```
# t-test Assumptions  
### Normality  
```{r, message=FALSE}
#***************************************************************
#ttest
  #alpha is 0.05
  #normality
qqnorm(height$growthrate_global)
qqline(height$growthrate_global)
#based on qqplot normality could be assumed 
shapiro.test(height$growthrate_global) #p value is 0.0006381
    #--> not normaly distributed
#independence -> can be assumed
#equality of variance 

#wilcoxon test for treatment
wilcox.test(height$growthrate_global~height$Treatment) #p value 0.0004546

#kruskal test for variety
kruskal.test(height$Variety, height$growthrate_global) #p value 0.4576 -> not significant
#kruskal test for interaction 
interT.V=interaction(height$Variety, height$growthrate_global)
kruskal.test(interT.V, height$growthrate_global) #p value 0.2333 -> not significant

#ttest when normality is assumed based on the qqplot
#equality of variance
bartlett.test(height$growthrate_global~height$Treatment) #equality of variance cannot be assumed


lm.growth=lm(height$growthrate_global~height$Treatment*height$Variety)
plot(lm.growth)
summary(lm.growth)
```
